### 2014-08-07 doc/thesis/typedlua.tex
Moving typing rules to the thesis and removing old documentation files
Rewriting the begining of the section about Typed Lua

Typed Lua allows statically typed Lua code to coexist and interact
with dynamically typed Lua code.
The Typed Lua compiler checks for type errors, but alwayas removes
the annotations, generating Lua code that runs in unmodified Lua
implementations.
The programmer can enjoy some of the benefits of static types even
without converting existing Lua modules to Typed Lua:
a dynamically typed module can export a statically typed interface,
and statically typed users of the module will have their use of the
module checked by the compiler.

Unlike gradual typing, Typed Lua does not insert run-time checks
between dynamically typed and statically typed parts of the program.
Unlike some optional type systems, the statically typed subset of
Typed Lua aims to be sound, so a later version of the Typed Lua
compiler can switch to gradual typing instead of just optional typing.

In this section we cover the design of Typed Lua:
In Section \ref{sec:types} we present the types that are available
in Typed Lua along with the subtyping and consistent-subtyping
relations.
In Section \ref{sec:examples} we present the design of Typed Lua
through code examples.
In Section \ref{sec:rules} we present the design of Typed Lua
through typing rules.

@@

The main goal of Typed Lua is to allow programmers to combine static
and dynamic typing in Lua, so Typed Lua is an extension to Lua 5.2
syntax that introduces optional type annotations, and class-based
object-oriented programming through the definition of classes,
interfaces, and modules.
An important feature of Typed Lua is its backwards compatibility with
Lua, that is, unannotated Typed Lua code is valid Lua code.
For this reason, the syntactic extensions introduced by Typed Lua
include new keywords that are not reserved words.
Appendix \ref{app:syntax} presents the complete syntax of Typed Lua,
which we cover in this section through code examples.

@@

Unlike gradual typing, we do not insert run-time checks that
inspect the interaction between dynamically typed code and statically
typed code during run-time.
In other words, Typed Lua does not guarantee that dynamically typed
code does not violate statically typed code.
Typed Lua does not insert run-time checks because they can decrease
run-time performance \citep{allende2013cis}.
We believe that a careful evaluation of run-time checks should be done
before inserting them on the type system.
In the previous example, we cannot guarantee that \texttt{dist} is never
going to call \texttt{abs} with a parameter that is not a number,
because in the semantics of Lua the minus operator can result in a
value that is not a number.
Still, we can implement run-time checks in the future because we are
using the consistent-subtyping relation of gradual typing to
formalize our optional type system, and the type system is being
designed to be sound.

@@

>>>>

Typed Lua is an optional type system for Lua, and its main goal is to
allow programmers to combine static and dynamic typing in Lua.
In practice, Typed Lua extends the syntax of Lua to introduce
optional type annotations and class-based object-oriented
programming through the definition of classes, interfaces, and
modules.
In addition, Typed Lua performs local type inference \citep{pierce2000lti}
to detect more precise types for unannotated expressions.
Even though the compiler warns the programmer about type errors,
it always removes the type annotations to generate Lua code that
runs in unmodified Lua implementations.

An important feature of Typed Lua is its backwards compatibility with Lua,
that is, unannotated Typed Lua code is valid Lua code.
For this reason, the syntactic extensions introduced by Typed Lua
include new keywords that are not reserved words.
Appendix \ref{app:syntax} presents the complete syntax of Typed Lua.

Another important feature of Typed Lua is its interface between
statically and dynamically typed modules. 
This means that the programmer can enjoy some of the benefits of
static types even without converting existing Lua modules to Typed Lua,
as a dynamically typed module can export a statically typed interface,
and statically typed users of the module have their use of the module
checked by the compiler.

Although we use a gradual type system to formalize our optional type system,
Typed Lua does not insert run-time checks in the gradual typing style.
In gradual typing, run-time checks inspect the interaction between
dynamically typed and statically typed code to guarantee that dynamically
typed code does not violate statically typed code during run-time.
A good reason for not inserting run-time checks at this moment is that they
can decrease run-time performance \citep{allende2013cis}, and we believe that
a careful evaluation of run-time checks should be done before inserting them
on the type system.

Unlike some optional type systems, we are designing Typed Lua aiming soundness
to make it possible to switch Typed Lua from optional typing to gradual typing
in the future, as a sound type system is the first step to make Typed Lua
insert run-time checks after static type checking.

In this section we show that Typed Lua is rich enough to preserve some Lua idioms.
First, we present the abstract syntax of Typed Lua types along with the
relationship among types.
Then we use some code examples to present an overview about Typed Lua and
its syntactical extensions.
After that we present the formalization of Typed Lua.

@@

In this section we use some examples of Typed Lua code to show how
they relate to Lua.
This examples shall give an informal overview of the type system.
It worth mentioning that we used our implementation to produce and
test all the examples that we present along this section.

@@

In the previous example, we cannot guarantee that \texttt{dist} is never
going to call \texttt{abs} with a parameter that is not a number,
because in the semantics of Lua the minus operator can result in a
value that is not a number.

@@

\begin{figure}[!ht]
\textbf{Abstract Syntax}\\
\dstart
$$
\begin{array}{rl}
s ::= & \;\; \mathbf{skip} \; | \;
s_{1} \; ; \; s_{2} \; | \;
\vec{l} = \vec{e}  \; | \;
\mathbf{while} \; e \; \mathbf{do} \; s \; | \;
\mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}\\
& | \; \mathbf{local} \; \vec{n}{:}\vec{T} = \vec{e} \; \mathbf{in} \; s \; | \;
\mathbf{rec} \; n{:}T = f \; \mathbf{in} \; s \; | \;
\mathbf{return} \; \vec{e} \; | \;
a\\
e ::= & \;\; \mathbf{nil} \; | \;
\mathbf{false} \; | \;
\mathbf{true} \; | \;
{<}{\it number}{>} \; | \;
{<}{\it string}{>} \; | \;
{...} \; | \;
f \; | \;
\{ \; \vec{c}\; \} \\
& | \; e_{1} + e_{2} \; | \;
e_{1} \; {..} \; e_{2} \; | \;
e_{1} == e_{2} \; | \;
e_{1} < e_{2} \; | \;
e_{1} \; \mathbf{and} \; e_{2} \; | \;
e_{1} \; \mathbf{or} \; e_{2} \\
& | \; \mathbf{not} \; e \; | \;
- e \; | \;
\# \; e \; | \;
a \; | \;
l \\
l ::= & \; n \; | \;
e_{1}[e_{2}]\\
a ::= & \;\; e_{1}(\vec{e_{2}}) \; | \;
e_{1}{:}n(\vec{e_{2}})\\
f ::= & \; \mathbf{fun} \; (){:}R \; s \; | \;
\mathbf{fun} \; ({...}{:}T){:}R \; s \; | \;
\mathbf{fun} \; (\vec{n}{:}\vec{T}){:}R \; s \; | \;
\mathbf{fun} \; (\vec{n}{:}\vec{T},{...}{:}T){:}R \; s\\
c ::= & \; [e_{1}] = e_{2}\\
n ::= & \; {<}{\it name}{>}\\
\end{array}
$$
\dend
\caption{Typed Lua abstract syntax}
\label{fig:syntax}
\end{figure}
